Online supplement accompanying `The Locally Gaussian Partial Correlation'

### 2020-09-02 lgpc_JBES_supp.tex
Implementing comments from Dag and one referee

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Online supplement accompanying `The Locally Gaussian Partial Correlation'},
  pdfauthor={Håkon Otneim and Dag Tjøstheim},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}

@@

\newcommand{\y}{\bm{y}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\hY}{\widehat{Y}}

@@

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density

\begin{equation}
\psi(\z) = \frac{1}{(2\pi)^{(d_1+d_2+d_3)/2}|\fSigma(\z)|^{1/2}} \exp \left\{-\frac{1}{2}(\z - \fmu(\z))^T\fSigma^{-1}(\z)(\z - \fmu(\z))\right\}
\label{eq:localgaussian0}
\end{equation}
and the locally Gaussian \emph{conditional} density

@@

Let \(\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}\). The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of \(\D^{-1/2}\fSigma_{12|3}\D^{-1/2}\), and so the LGPC matrix is thus given as the upper right hand block of the matrix
\begin{equation}
\D^{-1/2}\fSigma_{12|3}\D^{-1/2} = \D^{-1/2}\begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}\D^{-1/2}.
\label{eq:matrixdef}
\end{equation}
Let us consider this expression for a few special cases.

@@

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
\label{eq:penalty}
\end{equation}
where \(f\) is the density function of \((Z_i, Z_j)\). Note in particular that \(\rho_{0,b} \rightarrow \rho_0\) as \(b\rightarrow0\), which is the true local correlation parameter. The reader may consult Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for more details and regularity conditions in this regard. The corresponding estimate \(\rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}. Note that, writing \(\rho\) for \(\rho_b\),
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case. A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

@@

The second term in \(\Jb\) can similarly be written as \(II_J = B(\psi(\z, \theta_h(\z)) - f(\z)) + o(b^2)\), where \(B = \nabla \fv(\z, \theta_b(\z))\) are second derivatives of the local likelihood function with respect to the local correlations, and \(\psi(\z, \theta_b(\z)) - f(\z) = O(b^2)\) according to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}). Hence, it follows that the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is \(O(b^{-2})\), which again means that the convergence rate in eq. \eqref{eq:fixed-h} above must be modified accordingly in order to balance the convergence in eq. (20) in the main article. Extracting an analytic expression for the leading term of the asymptotic covariance matrix is, as mentioned by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), possible by means of symbolic manipulation software, but the expression is very complicated and of little practical use. See also the corresponding proof of Theorem 3 in Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}).

@@

Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables \((Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)\), where \(F_j\) is the cdf of \(X_j\) for \(i = 1,\ldots,p\), relies on proving asymptotic normality for the variables
\[Y_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),\]
where \(\fv(\cdot) = \partial L(\cdot)/\partial\frho\) in the same way as in Section \ref{app-loccor} in this supplement. This has been done by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) in the iid case using arguments from Schervish (\protect\hyperlink{ref-schervish1995theory}{2012}), and by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) in the \(\alpha\)-mixing case, using arguments from Fan and Yao (\protect\hyperlink{ref-fan2008nonlinear}{2008}). We need here to prove asymptotic normality of the variables

\[\hY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),\]
with \(\hnZ_t = \Phi^{-1}(F_n(\X_t))\), and \(F_n(\X_t)\) is the application of the marginal empirical distribution functions to the corresponding components of the vector \(\X_t\).

In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using \emph{pairs} \((i,j)\) of variables. It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion. Write \(\hY_n(\z) = Y_n(\z) - (Y_n(\z) - \hY_n(\z))\) and do a Taylor expansion of \(Y_n(\z)\) around \(F_n(\X_t)\). Since we use product kernels, we then need derivatives of (assuming without loss of generality that \(K(x) = K(-x)\))

@@

In the following we omit the index \(i=1,2\). We need a bound on the supremum of \(|F_n(X_t) - F(X_t)|\) and \(|F_n(x) - F(x)|\) as \(n \rightarrow \infty\). According to Arcones (\protect\hyperlink{ref-arcones1995law}{1995}), the best law of the iterated logarithm has been obtained by Rio (\protect\hyperlink{ref-rio1995functional}{1995}). One statement of this is as follows:

Let \(f(x) \in L_p\), \(2<p\leq\infty\), \(\sum\alpha_k k^{2/(p-2)} < \infty\), then the sequence \(f(X_t)\) satisfies the compact LIL (\emph{law of the iterated logarithms}), i.e.
\[\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}\]
is relatively compact and its limit set is \([-\sigma, \sigma]\) where
\[\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).\]
For a fixed \(x\), take \(f(X_t) = I(X_t\leq x)\). Then E\((f(X_t)) = F(x)\) and \(f(x)\in L_p\) for all \(p\), and
\[\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}\]
has its limit set in \([-\sigma, \sigma]\). Here
\[\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},\]
and according to Davydov's Lemma (see e.g.~Fan and Yao (\protect\hyperlink{ref-fan2008nonlinear}{2008}) p.~278),
\[\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,\]
with \(\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}\). We may for example take \(p=q=4\), leading to \(\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1\).

Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition \(\sum \alpha_k k^{2/(p-2)} < \infty\) is satisfied, and the bounds of \(-\sigma\) and \(\sigma\) can be made independent of \(x\).

Having
\[F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)\]
means that
\[n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))\]
has its limits in \([-\sigma, \sigma]\) or that the supremum of \(|F_n(x) - F(x)|\) is almost surely bounded by \(n^{-1/2}(2\log\log n)^{1/2}\).

This immediately implies that
\[\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)\]
is almost surely (and consequentially in probability) bounded by \(n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}\) which for all conceivable bandwidths is faster than the convergence rates of the CLT for \(Y_n(\z)\), which is \((nb^2)^{-\frac{1}{2}}\).

@@

We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below. First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\), possibly a multivariate time series. Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\). Note further that it suffices to consider the true pseudo-observations \(\{\Z_t\}\) instead of the empirical pseudo observations \(\{\hZ_t\}\) in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block bootstrap the entire vector series \(\{\Z_t\}\). This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above. Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for further details on this distinction. Bootstrapped quantities will be denoted with a star (\(*\)) in the following.

@@

Consider the general case \(\Z = (\Z_1, \Z_2, \Z_3)\) where each component is \(d_i\)-dimensional correspondingly, with \(i = 1,2,3\), and \(\Z\), naturally, is \(d_1+d_2+d_3\)-dimensional, and consider also the local partial correlation matrix of \((\Z_1,\Z_2|\Z_3)\). In alternative (iii) in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
\[
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
\]
This forms a \(d_1 \times d_2\) matrix \({\falpha}\). One can test for conditional independence between \(Z_{1,i}\) and \(Z_{2,j}\) given \(\Z_3\) by testing whether \(\alpha_{ij}=0\), or one can test for conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\) by testing whether \({\falpha} = \{\alpha_{ij}\} = {\bm{0}}\). The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for \({\falpha} = {\bm{0}}\) by using a matrix norm, for instance the Frobenius norm, where \(||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0\). A more convenient test statistic is \(\sum_{i,j} \alpha_{ij}^2\). To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
\[
T = \int_S h(\frho(\z)) \di F(\z)
\]
where \(S\) is an a area of integration, \(F\) is the cumulative distribution function of \(\Z\), and \(\frho = \frho(\z)\), as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from \(\Z\). The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \ref{chap:estimation} one considers. It is reasonable to express \(T\) as a function of \(\frho\) since all partial correlations can be expressed as a function of this vector (see e.g.~again equation \eqref{eq:offdiag3}). One possible choice of the function \(h\) is \(h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)\), which is a function of \(\frho\) since the \(\alpha\)-s are functions of \(\frho\). The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) is \(h(\theta(x))\), where \(\theta(x)\) is at most 5-dimensional and \(\x\) is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector \(\bb\) be fixed and let \(n \to \infty\), and then we let \(\bb = (b_1,b_2) = (b,b) \to \bm{0}\).
Accordingly we define
\[
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
\]
where \(\frho_b\) is the population quantity for a fixed bandwidth \(\bb\), and \(T_{n,b}\) by
\[
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
\]
where \(\frho_{b,n}(\z)\) is the estimate of \(\frho(\z)\) obtained by pairwise likelihood for each component of \(\frho(\z)\).

@@

Let us briefly demonstrate a practical implementation of the LGPC in another situation involving real data. As in the empirical example in the main article, we look at Granger causality,

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Online supplement accompanying `The Locally Gaussian Partial Correlation'},
  pdfauthor={Håkon Otneim and Dag Tjøstheim},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}

@@

\newcommand{\y}{\bm{y}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\hY}{\widehat{Y}}
\newcommand{\fY}{\bm{Y}}
\newcommand{\hfY}{\widehat{\bm{Y}}}

@@

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density \(\psi(\z,\fv)\) at the point \(\fv = \z\),

\begin{equation}
\psi(\z) = \frac{1}{(2\pi)^{(d_1+d_2+d_3)/2}|\fSigma(\z)|^{1/2}} \exp \left\{-\frac{1}{2}(\z - \fmu(\z))^T\fSigma^{-1}(\z)(\z - \fmu(\z))\right\}
\label{eq:localgaussian0}
\end{equation}
and the locally Gaussian \emph{conditional} quantity

@@

Let \(\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}\). The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of \(\D^{-1/2}\fSigma_{12|3}\D^{-1/2}\), and so the LGPC matrix is thus given as the lower left hand block of the matrix
\begin{equation}
\D^{-1/2}\fSigma_{12|3}\D^{-1/2} = \D^{-1/2}\begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}\D^{-1/2}.
\label{eq:matrixdef}
\end{equation}
Let us consider this expression for a few special cases.

@@

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
\label{eq:penalty}
\end{equation}
where \(f\) is the density function of \((Z_i, Z_j)\). Note in particular that \(\rho_{0,b} \rightarrow \rho_0\) as \(b\rightarrow0\), which is the true local correlation parameter. The reader may consult Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for more details and regularity conditions in this regard. The corresponding estimate \(\widehat \rho = \rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}. Note that, writing \(\rho\) for \(\rho_b\),
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2.1 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case. A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

@@

The second term in \(\Jb\) can similarly be written as \(II_J = B(\psi(\z, \frho_b(\z)) - f(\z)) + o(b^2)\), where \(B = \nabla \fv(\z, \frho_b(\z))\) are second derivatives of the local likelihood function with respect to the local correlations, and \(\psi(\z, \frho_b(\z)) - f(\z) = O(b^2)\) according to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}). Hence, it follows that the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is \(O(b^{-2})\), which again means that the convergence rate in eq. \eqref{eq:fixed-h} above must be modified accordingly in order to balance the convergence in eq. (20) in the main article. Extracting an analytic expression for the leading term of the asymptotic covariance matrix is, as mentioned by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), possible by means of symbolic manipulation software, but the expression is very complicated and of little practical use. See also the corresponding proof of Theorem 3 in Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}).

@@

Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables \((Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)\), where \(F_j\) is the cdf of \(X_j\) for \(i = 1,\ldots,p\), relies on proving asymptotic normality for the variables
\[\fY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),\]
where \(\fv(\cdot) = \partial L(\cdot)/\partial\frho\) in the same way as in Section \ref{app-loccor} in this supplement. This has been done by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) in the iid case using arguments from Schervish (\protect\hyperlink{ref-schervish1995theory}{2012}), and by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) in the \(\alpha\)-mixing case, using arguments from Fan and Yao (\protect\hyperlink{ref-fan2008nonlinear}{2008}). We need here to prove asymptotic normality of the variables

\[\hfY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),\]
with \(\hnZ_t = \Phi^{-1}(F_n(\X_t))\), and \(F_n(\X_t)\) is the application of the marginal empirical distribution functions to the corresponding components of the vector \(\X_t\).

In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using \emph{pairs} \((i,j)\) of variables. Then \(Y_n\) can be taken to be a scalar. It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion. Write \(\hY_n(\z) = Y_n(\z) - (Y_n(\z) - \hY_n(\z))\) and do a Taylor expansion of \(Y_n(\z)\) around \(F_n(\X_t)\). Since we use product kernels, we then need derivatives of (assuming without loss of generality that \(K(x) = K(-x)\))

@@

In the original version of this proof we used an argument based on the law of the iterated logarithm of Rio (\protect\hyperlink{ref-rio1995functional}{1995}), leading to a rate of \(n^{-1/2}(2 \log \log n )^{1/2}\), but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process. In the following we omit the index \(i=1,2\):

\begin{align*}
\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right).
\end{align*}

@@

We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below. First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\). Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\). Note further that it suffices to consider the true pseudo-observations \(\{\Z_t\}\) instead of the empirical pseudo observations \(\{\hZ_t\}\) in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block-bootstrap the entire vector series \(\{\Z_t\}\). This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above. Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for further details on this distinction. Bootstrapped quantities will be denoted with a star (\(*\)) in the following.

@@

Consider the general case \(\Z = (\Z_1, \Z_2, \Z_3)\) where each component is \(d_i\)-dimensional correspondingly, with \(i = 1,2,3\), and \(\Z\), naturally, is \(d_1+d_2+d_3\)-dimensional, and consider also the local partial correlation matrix of \((\Z_1,\Z_2|\Z_3)\). In alternative (iii), where \(d_3 = 1\), in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
\[
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
\]
This forms a \(d_1 \times d_2\) matrix \({\falpha}\). One can test for conditional independence between \(Z_{1,i}\) and \(Z_{2,j}\) given \(\Z_3\) by testing whether \(\alpha_{ij}=0\), or one can test for conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\) by testing whether \({\falpha} = \{\alpha_{ij}\} = {\bm{0}}\). The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for \({\falpha} = {\bm{0}}\) by using a matrix norm, for instance the Frobenius norm, where \(||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0\). A more convenient test statistic is \(\sum_{i,j} \alpha_{ij}^2\). To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
\[
T = \int_S h(\frho(\z)) \di F(\z)
\]
where \(S\) is an a area of integration, \(F\) is the cumulative distribution function of \(\Z\), and \(\frho = \frho(\z)\), as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from \(\Z\). The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \ref{chap:estimation} one considers. It is reasonable to express \(T\) as a function of \(\frho\) since all partial correlations can be expressed as a function of this vector (see e.g.~again equation \eqref{eq:offdiag3}). One possible choice of the function \(h\) is \(h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)\), which is a function of \(\frho\) since the \(\alpha\)-s are functions of \(\frho\). The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) is \(h(\theta(x))\), where \(\theta(x)\) is at most 5-dimensional and \(\x\) is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector \(\bb\) be fixed and let \(n \to \infty\), and then we let \(\bb = (b_1,b_2) = (b,b) \to \bm{0}\).
Accordingly we define
\[
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
\]
where \(\frho_b\) is the population quantity for a fixed bandwidth \(\bb\), and \(T_{n,b}\) by
\[
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
\]
where \(\frho_{b,n}(\z)\) is the estimate of \(\frho(\z)\) obtained by pairwise likelihood for each component of \(\frho(\z)\).

@@

@@

Let us briefly demonstrate a practical implementation of the LGPC in another situation involving real data. As in the empirical example in the main article, we look at Granger causality.
